---
title: "Project 1"
author: "Ifeoma Ojialor"
date: "9/15/2020"
output: 
  rmarkdown::github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Here are the required libraries needed for this project.

```{r packages, message=FALSE}
# required packages 
library(httr)
library(jsonlite)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(rmarkdown)
```

## Extracting and Accessing the NHL records API.

In this section, I will create the functions needed to contact the NHL records API and return well parsed data based on their endpoints.
Below is the function needed to access the franchise endpoint. This function should return the id, firstSeasonId, lastSeasonId, and the name of every team in the history of the NHL.

```{r get_franchise}
get_fran <- function(tab_name, ID=NULL){
  base_url <- "https://records.nhl.com/site/api"
  full_url <- paste0(base_url, "/", tab_name)
  get_url <- GET(full_url, ID)
  txt_url <- content(get_url, "text")
  franc_data <- fromJSON(txt_url, flatten=TRUE)
  franc_data <- as.data.frame(franc_data)
  if (!is.null(ID)){
    franc_data <- filter(franc_data, ID==ID)
  }
  return(franc_data)
}
```

We also need to create functions to access each of the following endpoints. Below are the endpoints we need to access in this section.

+ /franchise-team-totals
     
+ /franchise-season-records?cayenneExp=franchiseId=ID  

+ /franchise-goalie-records?cayenneExp=franchiseId=ID 

+ /franchise-skater-records?cayenneExp=franchiseId=ID 

```{r get_franc_totals}
get_fran_tot <- function(tab_name, ID){
  if (is.null(ID)){
    stop("Need a valid franchise ID")
  }
  if(!is.numeric(ID)){
    stop("Franchise ID must be a number")
  }
  base_url <- "https://records.nhl.com/site/api"
  full_url <- paste0(base_url, "/", tab_name)
  get_url <- GET(full_url)
  txt_url <- content(get_url, "text")
  ftotal_data <- fromJSON(txt_url, flatten=TRUE)
  ftotal_data <- as.data.frame(ftotal_data)
  return(ftotal_data)
}
```

```{r }
get_fran_rec <- function(tab_name, ID){
  if (is.null(ID)){
    stop("Need a valid franchise ID")
  }
  base_url <- "https://records.nhl.com/site/api"
  f_path <- "cayenneExp=franchiseId="
  full_url <- paste0(base_url, "/", tab_name, "?",f_path,as.character(ID))
  get_url <- GET(full_url)
  txt_url <- content(get_url, "text")
  goali_data <- fromJSON(txt_url, flatten=TRUE)
  goali_data <- as.data.frame(goali_data)
  return(goali_data)
}
```


```{r }
get_goalie <- function(tab_name, ID){
  if (is.null(ID)){
    stop("Need a valid franchise ID")
  }
  base_url <- "https://records.nhl.com/site/api"
  f_path <- "cayenneExp=franchiseId="
  full_url <- paste0(base_url, "/", tab_name, "?",f_path,as.character(ID))
  get_url <- GET(full_url)
  txt_url <- content(get_url, "text")
  goali_data <- fromJSON(txt_url, flatten=TRUE)
  goali_data <- as.data.frame(goali_data)
  return(goali_data)
}
```

```{r }
get_skater <- function(tab_name, ID){
  if (is.null(ID)){
    stop("Need a valid franchise ID")
  }
  base_url <- "https://records.nhl.com/site/api"
  f_path <- "cayenneExp=franchiseId="
  full_url <- paste0(base_url, "/", tab_name, "?",f_path,as.character(ID))
  get_url <- GET(full_url)
  txt_url <- content(get_url, "text")
  skater_data <- fromJSON(txt_url, flatten=TRUE)
  skater_data <- as.data.frame(skater_data)
  return(skater_data)
}
```


## Extracting and accessing the NHL stats API.
Above, we showed how to access the records API. Now, we will show how to access the NHL stats API and extract information from the teams endpoint using modifiers. Information about the eight modifiers are listed below;

+ ?expand=team.roster 
    * This modifier shows roster of active players for the specified team.

+ ?expand=person.names
    * This modifier shows the same as above, but gives less info.
    
+ ?expand=team.schedule.next 
    * This modifier shows the returns details of the upcoming game for a team.

+ ?expand=team.schedule.previous 
    * This modifier shows the same as above but for the last game played.
    
+ ?expand=team.stats 
    * This modifier returns the teams stats for the season.
    
+ ?expand=team.roster&season=20142015 
    * This modifier shows the roster for that season
+ ?teamId=4,5,29 
    * This modifier can string team id together to get multiple teams
    
+ ?stats=statsSingleSeasonPlayoffs 
    * This modifier specifies which stats to get. 
    
    
```{r get_stats_function}   
get_stats <- function(tab_name, ID=NULL, mod=NULL){
  base_url <- "https://statsapi.web.nhl.com/api/v1/"
  if (is.null(ID)){
    full_url <- paste0(base_url, "/", tab_name)}
    else {
      full_url <- paste0(base_url, "/", tab_name, "/",ID)}
  if (is.null(mod)){
    full_url <- paste0(base_url, "/", tab_name)}
  get_url <- GET(full_url)
  txt_url <- content(get_url, "text")
  stats_data <- fromJSON(txt_url, flatten=TRUE)
  stats_data <- as.data.frame(stats_data)
  return(stats_data)
}    
```    
   
Now we will create a wrapper function to make it easier to access any of the API's done above.

```{r wrap_fun}
get_data <- function(endpoint,ID,mod){
if(is.null(endpoint)){stop("Need valid endpoint")}
    else if (endpoint=="franchise_data"){get_franchise_data(ID)}
    else if (endpoint=="franchise_totals"){get_franchise_totals(ID)}
    else if (endpoint=="franchise_records"){get_franchise_records(ID)}
    else if (endpoint=="goalie_records"){get_goalie_records(ID)}
    else if (endpoint=="skater_records"){get_skater_records(ID)} 
    else if (endpoint=="team_stats"){get_stats(ID)} 

}

```
   
## Exploratory Data Analysis.
Now that we have our data, it is time to make use of that data. In this section, we will explore our data using plots and summaries.


Suppose we believe that Los Angeles is the country with the most wins. We can investigate this idea using our data. Let us start with the franchise data table. We can see that the data involves a span of 100 years(1917 to 2017). Lets investigate whether the amount of games played every 50 years starting from year 1917 increased or decreased.

```{r}
# We can access this data using our get_fran function
fran <- get_fran(tab_name = "franchise")

# Now lets perform the necessary transformation and visualize the data
teamData <- separate(fran,data.firstSeasonId,into=c("startYear","Quarter"),sep=4)
teamData <- teamData %>% mutate(Year = "NA") 
teamData$Year <- ifelse(teamData$startYear <= 1967, 1,0)
teamData$Year <- as.factor(teamData$Year)
g <- ggplot(teamData, aes(x = as.numeric(startYear)))
g+geom_histogram(aes(fill = Year), position = "dodge") +xlab("Franchise Start Year") +scale_fill_discrete(labels = c("1968-2017", "1917-1967"))   +labs(title = "Histogram of Start Year")
```

From the plot above, we see that more games were played during 1968 - 2017 and less games were played during 1917-1967. Now, lets investigate the relationship between year and number of games played.



```{r, message=FALSE}

#Join by franchiseId so that totalGames column is included
d <- get_fran_tot(tab_name = "franchise-team-totals", ID=14)
e<-d %>% inner_join(teamData, by = "data.id")

#create scatter plot of franchise year vs.total games played
y <- filter(e, startYear %in% (1917:1967),data.gamesPlayed > 2000)
g<-ggplot(y,aes(x=startYear,y=data.gamesPlayed))
g+geom_point()+ylab("Total Games Played")+xlab("Franchise Year")+ 
  geom_smooth(method=lm, col="Green") + 
  ggtitle("Total Games Played by Franchise Year")


z <- filter(e, startYear %in% (1968:2017),data.gamesPlayed > 2000)
g<-ggplot(z,aes(x=startYear,y=data.gamesPlayed))

g+geom_point()+ylab("Total Games Played")+xlab("Franchise Year")+ 
  geom_smooth(method=lm, col="Green") + 
  ggtitle("Total Games Played by Franchise Year")
```
From the scatterplots above we can infer that there is no linear relationship between the year and the number of games played. This could be because the games where played by teams who were not active. It is somewhat intuitive that there is no relationship between the two variables because some teams stopped playing as the year passed by. Lets investigate this hypothesis with the use of a boxplot. 

```{r }
rangers <- get_fran_tot(tab_name = "franchise-team-totals", ID=14)

#Now lets check which is the winning team, we will investigate the team whose wins are greater than losses.
rangers <- mutate(rangers, net_wins = as.numeric(rangers$data.wins - rangers$data.losses))
rangers <- mutate(rangers, net_wins_stats = as.numeric(rangers$data.wins - rangers$data.losses)<0)
rangers <- mutate(rangers, active=is.na(data.lastSeasonId))

g<-ggplot(rangers,aes(x=active))
g+geom_boxplot(aes(y=data.gamesPlayed,fill=active))+theme(legend.position="none")+xlab("Is Franchise Still Active?")+
  ylab("Total Games Played")+labs(title="Total Games by Active Status")+scale_x_discrete(labels=c("No","Yes"))


```
 
With this boxplot, we can see that the teams that were active played a higher number of games than teams that were not active. Recall that were are interested in the total number of games played by the rangers team. Lets investigate this further by creating a contingency table of the wins and losses by a Los Angeles team in 1967.

```{r }

#Lets visualize the teams active.
ggplot(rangers, aes(x=active)) + geom_bar(colour = "red") + xlab("Number of teams active")
#Let us create a contingency table of the Wins and losses of each team.
rangers$net_wins_stats<-str_replace_all(rangers$net_wins_stats, 'TRUE', 'Losing Team')
rangers$net_wins_stats<-str_replace_all(rangers$net_wins_stats, 'FALSE', 'Winning Team')

#maxW <- rangers %>% group_by(data.teamName) %>% summarise(mean(net_wins_stats))
tbl <- table(rangers$data.teamName,rangers$net_wins_stats)
knitr::kable(tbl)
```

From above, we see that more teams are active , we see that we have a couple of teams who won a game atleast twice. Lets investigate this further by looking at teams who are active and won a game atleast twice.

```{r}
# Lets investigate the winning teams that are active  

rangers2 <- rangers %>% mutate(active=is.na(data.lastSeasonId)) %>% group_by(data.teamName)
table(rangers2$data.teamName, rangers2$active)
```

We can see that some teams are active while some are not. Lets obtain quantitative summaries of number of games played for each team.

```{r}
stats<-rangers2 %>% group_by(active, data.teamName) %>% summarise(averageGames=mean(data.gamesPlayed), medGame= median(data.gamesPlayed), maxGame = max(data.gamesPlayed)) %>% arrange(desc(maxGame))
stats



```

Alternatively, we can create boxplots;
```{r}
g<-ggplot(rangers2,aes(x=active))
g+geom_boxplot(aes(y=data.gamesPlayed,fill=active))+theme(legend.position="none")+xlab("Is Franchise Still Active?")+
  ylab("Total Games Played")+labs(title="Total Games by Active Status")+scale_x_discrete(labels=c("No","Yes"))
```
Based on the quantitative summaries, we see that Montreal Canadiens from the rangers played the most games.

